---
title: "analysis_code"
author: "Chiyoko Onouye"
date: "2023-03-31"
output: html_document
---

# Analysis script

# 

#this script loads the cleaned data and generated graphs and plots to help visualize #the data to help anwser questions and saves it as Rds file in the Analysis_Code folder

## ---- packages --------

This needs to run to load the R script so we can include the code chunks below.

Note that you can pull in more than one R script if you want.

#load needed packages.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('Anaylsis_Penguins.R')
```

```{r, packages, message = FALSE, warning = FALSE}

```

## ---- loaddata --------

#load data.

```{r, loaddata}

```

#look at data

```{r, exploredata}

```

##1.Is there a difference between females and males among the three species of penguins?

```{r, penguinsex}


```

###Looking at just Adelie Sex comparison

```{r, AdelieSex}


```

###Looking at just Gentoo Sex comparison

```{r, GentooSex}


```

###Looking at just Chinstrap Sex comparison

```{r, ChinstrapSex}


```

##2. Are the 3 species of penguins different?

```{r, PenguinSpecies}


```

##3. Is there a difference in penguins between the regions

```{r, PenguinIslands}


```

## ---- savedata --------

```{r, finalizedata}
```

# location to save file

```{r, savedata}
```

